The typed monomorphic first-order formulas subset, called TFF0, of the TPTP language~\cite{TPTP}, is a representation language for many-sorted first-order logic. It contains \verb'if'-\verb'then'-\verb'else' and \verb'let'-\verb'in' constructs (see below), which is useful for applications, but is inconsistent in its treatment of the boolean sort. It has a predefined atomic sort symbol \verb'$o' denoting the boolean sort. However, unlike all other sort symbols, \verb|$o| can only be used to declare the return type of predicate symbols. This means that one cannot define a function having a boolean argument, use boolean variables or equality between booleans. 

Such an inconsistent use of the boolean sort results in having two kinds of \verb'if'-\verb'then'-\verb'else' expressions and four kinds of \verb'let'-\verb'in' expressions. For example, a \folb-term $\letin{f(x_1:\sigma_1, \ldots, x_n:\sigma_n)}{s}{t}$ can be represented using one of the four TPTP alternatives \verb|$let_tt|, \verb|$let_tf|, \verb|$let_ft| or \verb|$let_ff|, depending on whether $s$ and $t$ are terms or formulas. 

Since the boolean type is second-class in TPTP, one cannot directly represent formulas coming from program analysis and interactive theorem provers, such as formulas \eqref{formula:contains} and \eqref{formula:subset-sorted} of Section~\ref{sec:cicm15/introduction}.

We propose to modify the TFF0 language of TPTP to coincide with \folb. It is not late to do so, since there is no general support for \verb'if'-\verb'then'-\verb'else' and \verb'let'-\verb'in'. To the best of our knowledge, Vampire is currently the only theorem prover supporting full TFF0. Note that such a modification of TPTP would make multiple forms of \verb'if'-\verb'then'-\verb'else' and \verb'let'-\verb'in' redundant. It will also make it possible to directly represent the SMT-LIB core theory.

We note that our changes and modifications on TFF0 can also be applied to the TFF1 language of TPTP~\cite{tff1}. TFF1 is  a polymorphic extension of TFF0 and its formalisation  does not treat the boolean sort. Extending our work to TFF1 should not be hard but has to be done in detail.