% !TEX root = main.tex

%\noindent The work described in this thesis has been carried out in six papers, each contained in a separate chapter.

{\setstretch{1.075}

\noindent This thesis is based on the work contained in the following papers, each presented in a separate chapter.

\noindent\paragraph{\hyperref[chap:fool]{Chapter~1.}} A First Class Boolean Sort in First-Order Theorem Proving and TPTP. \emph{Evgenii Kotelnikov, Laura Kov\'acs and Andrei Voronkov.} In \emph{Proceedings of the 8th Conference on Intelligent Computer Mathematics}, pages 71--86. Springer, 2015.

\noindent\paragraph{\hyperref[chap:implementation]{Chapter~2.}} The Vampire and the FOOL. \emph{Evgenii Kotelnikov, Laura Kov\'{a}cs, Giles Reger and Andrei Voronkov.} In \emph{Proceedings of the 5th ACM SIGPLAN Conference on Certified Programs and Proofs}, pages 37--48. ACM New York, 2016.

\noindent\paragraph{\hyperref[chap:cnf]{Chapter~3.}} A Clausal Normal Form Translation for FOOL. \emph{Evgenii Kotelnikov, Laura Kov\'{a}cs, Martin Suda and Andrei Voronkov.} In the \emph{Proceedings of the 2nd Global Conference on Artificial Intelligence}, pages 53--71. EPiC Series in Computing, 2016.

\noindent\paragraph{\hyperref[chap:boogie]{Chapter~4.}} A FOOLish Encoding of the Next State Relations of Imperative Programs. \emph{Evgenii Kotelnikov, Laura Kov\'{a}cs and Andrei Voronkov.} In \emph{Proceedings of the 9th International Joint Conference on Automated Reasoning}, pages 405--421. Springer, 2018.

\noindent\paragraph{\hyperref[chap:aws]{Chapter~5.}} Checking Network Reachability Properties by Automated Reasoning in First-Order Logic. \emph{Evgenii Kotelnikov and Pavle Suboti\'{c}.} A technical report based on a joint work with Byron~Cook, Temesghen Kahsai and Sean~McLaughlin, 2018.

\noindent\paragraph{\hyperref[chap:tfx]{Chapter~6.}} TFX: The TPTP Extended Typed First-Order Form. \emph{Geoff Sutcliffe and Evgenii Kotelnikov.} In \emph{Proceedings of the 6th Workshop on Practical Aspects of Automated Reasoning}, pages 72--87. CEUR Workshop Proceedings, 2018.

}